A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-15 code/table1.Rmd
move all code blocks to child Rmd files

```{r table-1}
load("../data/cluster_spec_enrich_table1.Rda")
cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```

>>>>

```{r table-1}
load("../data/cluster_spec_enrich_table1.Rda")
cluster_spec_enrich_table1 %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>

### 2020-07-15 text/manuscript.Rmd
move all code blocks to child Rmd files

After selecting our pathways and genes, we then compiled a matrix of the
pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
member gene carries a mutation) to the pathway regardless of the number of genes
carrying mutations (this data is visualized in **figure 1a**).
```{r child='data_set_generation.Rmd'}
```

```{r child='figure_1.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 1:
Classification of tumors by pathway disruptions.** a) Each of 377 selected
Reactome pathways (columns) is classified as disrupted if one or more genes is
mutated in the tumor sample (rows). Tumors are grouped by tissue of origin using
standardized abrreviations from the TCGA project. b) MCA analysis was performed
restricted to the 36 informative components. c) The resulting MCA was projected
down to 3 dimensions for the purpose of density based clustering (only the first
2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was used to
perform hierarchical density based clustering tuned to minimize the fraction of
unassigned tumors. e) HDBSCAN clustered tumors were visualized on the UMAP
projection; tumors are colored by cluster membership (see legend). This results
in some tumors remaining unclassified (grey points). f) kNN was used to assign
membership to the remaining tumors. g) Violin plots of sample mutation counts
for tumors in each class. h) Projection of cancer type identity onto UMAP.
Abbrevs: BRCA = Breast cancer, COAD = Colorectal adenocarcinoma, HNSC = Head and
Neck Squamous Cell, PAAD = Pancreatic adenocarcinoma. i) Projection of breast
cancer subtypes onto the UMAP. j) projection of BRCA1/2 somatic mutation onto the
UMAP. k) projection of tumor stage onto the UMAP,
regardless of cancer type. l) Projection of metastatic status onto the UMAP.
Abbrevs: M0 = non-metastatic tumors, M1 = metastatic tumors.</p>

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
replacing genes with pathways (**figure 2**). A complete list of pathway
disruptions with percent mutated samples and top genes is provided in
**supplementary table ST1**). These reveal the broad importance of many well
known pathways that are disrupted in cancer, including <mark>PI3K/Akt
activation, DNA repair etc fix this with top 5 or so quoted pathway
labels</mark>. The top third of pathways drop off gradually in overall mutation
frequency, possibly owing to relative diversity among all tumors.
```{r child='figure_2.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;"><b>Figure 2:
Pan-cancer enrichment of pathway disruptions.</b> "Oncoprint" style graphic in
which each column represents a tumor and each row represents a pathway. The
pathways are ranked such that the pathway with the most disruptions across all
7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
top pathway disruptions successively. All 377 pathways are shown. Red signifies
pathway disrupted pathways, grey signifies not disrupted.</p>

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
0 pathways in class 8 (for which most pathways are putatively disrupted), 25
pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
had enrichment. To explore the specific pathways marking each class, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
metabolic pathways including RNA and protein biosynthesis (**supplementary
figure S4**). Most class 4 enriched pathways were not unique to this class but
shared in class 10 & 12, suggesting that this grouping reflects either greater
heterogeneity or possibly a lower stage of tumor evolution (as distinct from
clinical ``stage''). Classes 9-12 have in common mutations in extracellular
signaling, intracellular signaling, and immune-related pathways (see **figures
3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
distinct of the 12, we chose to investigate these two groups further.
```{r child='figure_3.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3. Pathway-specific enrichment within clusters.** Heatmap shows relative enrichment of each pathway (rows) within each numbered cluster (columns). The annotation at left indicates the Reactome pathway group at the second level (check this).</p>

@@

In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
the highest pathway enrichment scores of the three major categories, though
there is a significant number of tumors spatially distributed between cluster
groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
pre-replicative complex," "metabolism of polyamines," "degradation of
$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
with aspects of hedgehog signaling. To determine once again whether the genes
driving pathway enrichment were commonly shared among these pathways we plotted
and performed unsupervised clustering on tumors with an individual mutation
matrix using every gene in each of the ten pathways, grouped by shared pathway
membership (**figure 4**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
genes in the proteosomal pathway whose activity
contributes to all the aforementioned pathways and the majority of tumors in
these classes have a single mutation in one of these genes. Nonetheless, a
significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
proteasomal mutations but instead harbor mutations in genes representing a
majority of the same pathways, suggesting that most if not all of the pathways
are individually important and that proteasomal mutations may be sufficient to
drive the net effect of the other mutations in this cluster. We annotated the
tumors in this class according to tissue of origin to determine whether
proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
indeed we could observe some clustering. These data could explain how
tissue-specific gene mutation signatures might result in general cancer
phenotypes by converging at the pathway level.
```{r child='figure_4.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey). The genes are organized according to which of the top distinguishing pathways for this class that they effect. Some genes affect multiple pathways, as indicated by the bars at left. The highlighted genes at right are all proteasomal subunits.</p>

@@

<mark>As discussed previously, prior efforts to extract common signatures from
pan-cancer datasets met with difficulty in distinguishing tumor samples from
their tissue-specific -omics data signatures. Given our pathway-disruption based
classification, this raises the question, are tumor phenotypes driven entirely
by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
few samples to detect above statistical significance thresholds), or a
combination of both? To answer this question, we ranked the top cluster-specific
pathways for each cluster, and then tabulated the most frequently mutated genes
for each pathway as a function of cancer type (**Table 1** and **Supplementary
tables ST2-ST13**). We find that in most clusters the top pathways are
determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
However in many cancers we find divergence outside these top driver mutations,
wherein etc etc</mark>
```{r child='table_1.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>

@@

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found 99 enriched pathways across all clusters (**table 2**). A number of
enrichments were found in multiple clusters represented by pathways that were
already shown to be enriched in a neighboring cluster. For example,
“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
Cluster 1, one of two clusters for which we could not identify any
distinguishing pathways (cluster 8 being the other), had enrichment in two
pathways that were otherwise specific to non-metastatic tumors of other
clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
samples were enriched for two key neutrophil pathways, “Neutrophil
Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.
```{r child='table_2.Rmd'}
```

@@

### Supplementary Figure S1
```{r child = 'figure_s1.Rmd'}
```
![**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]

### Supplementary Figure S2
```{r child = 'figure_s2.Rmd'}
```

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S3
```{r child = 'figure_s3.Rmd'}
```
![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S4
```{r child = 'figure_s4.Rmd'}
```

![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]

### Supplementary Table ST1
```{r child = 'table_s1.Rmd'}
```

![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]

![**Supplemental figure S6. Survival fit independent of disease status.** _The red and blue curves show the survival of people in the US by female and male, respectively. The orange and green curves show the model fits._](../figures/fig6A.png)

>>>>

After selecting our pathways and genes, we then compiled a matrix of the
pathways, assigning a Boolean value (a 1 or 0, indicating whether any pathway
member gene carries a mutation) to the pathway regardless of the number of genes
carrying mutations (this data is visualized in **figure 1a**).

```{r child='../code/figure1.Rmd'}
```

@@

In order to visualize pathway disruption enrichment across all cancers
(pan-cancer), we created oncoprint graphs analagous to the commonly use method,
replacing genes with pathways (**figure 2**). A complete list of pathway
disruptions with percent mutated samples and top genes is provided in
**supplementary table ST1**). These reveal the broad importance of many well
known pathways that are disrupted in cancer, including <mark>PI3K/Akt
activation, DNA repair etc fix this with top 5 or so quoted pathway
labels</mark>. The top third of pathways drop off gradually in overall mutation
frequency, possibly owing to relative diversity among all tumors.

```{r child='../code/figure2.Rmd'}
```

To discover what pathways drive the UMAP clustering, we calculated enrichment
for each pathway _in-class_ relative to all other classes combined (see methods)
and ranked pathways for each class from highest to lowest enrichment. We
visualized the enrichment as a heatmap (**figure 3**). Using this approach, we
identified 3 pathways highly enriched (enrichment score > 0.2, see methods) in
class 1, 11 pathways enriched in class 2, 9 pathways in class 3, 12 pathways in
class 4, 23 pathways in class 5, 31 pathways in class 6, 35 pathways in class 7,
0 pathways in class 8 (for which most pathways are putatively disrupted), 25
pathways in class 9, 28 pathways in class 10, 24 pathways in class 11, and 29
pathways in class 12. Some pairs of classes, in particular classes 6 & 7,
classes 9 & 10, and 11 & 12 were highly concordant in terms of which pathways
had enrichment. To explore the specific pathways marking each class, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). <mark>Class x</mark> was distinguished by
metabolic pathways including RNA and protein biosynthesis (**supplementary
figure S4**). Most class 4 enriched pathways were not unique to this class but
shared in class 10 & 12, suggesting that this grouping reflects either greater
heterogeneity or possibly a lower stage of tumor evolution (as distinct from
clinical ``stage''). Classes 9-12 have in common mutations in extracellular
signaling, intracellular signaling, and immune-related pathways (see **figures
3** and **supplementary figure S4**). Cluster group 6-7 had the highest pathway
enrichment levels of the three, having mutations in hedgehog signaling,
$\beta$-catenin degradation, cellular response to hypoxia, regulation of cell
cycle and apoptosis among others. As cluster groups 9-12 and 6-7 were the most
distinct of the 12, we chose to investigate these two groups further.
<div style="text-align: right"> Figure 3 data </div>

```{r child='../code/figure3.Rmd'}
```

@@

In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
the highest pathway enrichment scores of the three major categories, though
there is a significant number of tumors spatially distributed between cluster
groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
pre-replicative complex," "metabolism of polyamines," "degradation of
$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
with aspects of hedgehog signaling. To determine once again whether the genes
driving pathway enrichment were commonly shared among these pathways we plotted
and performed unsupervised clustering on tumors with an individual mutation
matrix using every gene in each of the ten pathways, grouped by shared pathway
membership (**figure 4**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
genes in the proteosomal pathway whose activity
contributes to all the aforementioned pathways and the majority of tumors in
these classes have a single mutation in one of these genes. Nonetheless, a
significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
proteasomal mutations but instead harbor mutations in genes representing a
majority of the same pathways, suggesting that most if not all of the pathways
are individually important and that proteasomal mutations may be sufficient to
drive the net effect of the other mutations in this cluster. We annotated the
tumors in this class according to tissue of origin to determine whether
proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
indeed we could observe some clustering. These data could explain how
tissue-specific gene mutation signatures might result in general cancer
phenotypes by converging at the pathway level.

```{r child='../code/figure4.Rmd'}
```

@@

<mark>As discussed previously, prior efforts to extract common signatures from
pan-cancer datasets met with difficulty in distinguishing tumor samples from
their tissue-specific -omics data signatures. Given our pathway-disruption based
classification, this raises the question, are tumor phenotypes driven entirely
by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
few samples to detect above statistical significance thresholds), or a
combination of both? To answer this question, we ranked the top cluster-specific
pathways for each cluster, and then tabulated the most frequently mutated genes
for each pathway as a function of cancer type (**Table 1** and **Supplementary tables ST2-ST13**). 
We find that in most clusters the top pathways are
determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
However in many cancers we find divergence outside these top driver mutations,
wherein etc etc</mark>

```{r child='../code/table1.Rmd'}
```

@@

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found 99 enriched pathways across all clusters (**table 2**). A number of
enrichments were found in multiple clusters represented by pathways that were
already shown to be enriched in a neighboring cluster. For example,
“Beta-catenin independent WNT signaling” is enriched in non-metastatic samples
of clusters 6, 7, 11 and 12 (see **supplementary figure S4**) but not in 9 and
10. This pathway is enriched in metastatic tumors of clusters 9 and 10, plus the
union of 9+10 (p < 10-3, **table 2**). This is also true of “Erythropoietin
activates RAS”, which is enriched in clusters 11 and 12 (**supplementary figure
S4**) and also in metastatic tumors of clusters 9 and 10, plus cluster 7.
Cluster 1, one of two clusters for which we could not identify any
distinguishing pathways (cluster 8 being the other), had enrichment in two
pathways that were otherwise specific to non-metastatic tumors of other
clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7). Cluster 5 metastatic
samples were enriched for two key neutrophil pathways, “Neutrophil
Degranulation” and “Fc epsilon receptor (FCERI) signaling” which are specific to
cluster 9 & 10 non-metastatic tumors. Thus, pathway lesions specific to
metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.

```{r child='table2.Rmd'}
```

@@

```{r child='../code/supplements.Rmd'}
```
